---
title: "Lab 4: Factors"
author: "Environmental Data Analytics | John Fay and Luana Lima "
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---


```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
#Read packages
library(tidyverse);library(lubridate); library(here)

#Read in data, don't convert strings to factors
sites <- read.csv(here('Data/Raw/NWIS_SiteFlowData_NE_RAW.csv'))
dim(sites)

#Convert date column to date object
sites$date <- ymd_hms(sites$date) #now this is dates and times. year month day, hour minute seconds.that's what makes it format as POSIXct or POSIXt

```


## Examine the lakename column as a character column
* Tip - Lubridate cheatsheet: <https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf>
```{r create and explore days of week column}

#Compute the day of the week each sample was collected
sites$dow <- wday(sites$date) #wday is the numerical day of the year, and will return the day of the week.

## What class is the dow column?
class(sites$dow)

#List the unique values in the column
unique(sites$dow)

#Summary - is it meaningful?
summary(sites$dow)#this shows like the mean and stuff. unhelpful. not all numbers have numeric meaning. ordinal numbers of days of the week don't really hold meaning for like mean or median. they are labels, and shouldn't be numeric.should be a factor now

```

## Convert the days of the week to a factor
```{r convert dow column to factor and explore}

#Convert to a factor
sites$dow <- factor(sites$dow)

## What class is the dow column now?
class(sites$dow)

#List the unique values in the column
unique(sites$dow)

#Summary - is it meaningful?
summary(sites$dow) #shows them as categories. numbers, and then they're frequency in the dataset?

#Show the levels associated with our factor column
levels(sites$dow) #levels are. factors are a grouping variable. every row in the data set that has a dow column value of 4, they all have a factor in common. factor is a grouping variable. strings as factors, most of the time they are grouping variable. a string would NOT be a grouping variable if like if it was a DNA sequence. where the letters mean something, they are not a grouping variable.
```

## Factors & Levels
A **factor** is a grouping variable: rows with the same value in the factor column are members of the same group. For example, all rows with a `dow` of 3 were collected on a Tuesday.

**Levels** are the unique values that the factors can take. The `unique()` function reveals the levels as does the `levels()` function. By default, the levels includes only the values in the factor column. However, we can add additional levels and we can also re-order levels. Why would we want to do this? Well, one example would be if we wanted to plot the number of samples recorded each day of the week, and in that plot we wanted to include all days of the week, not just the one's found in our data...

```{r plot samples recorded each day of the week}
#Create a bar plot showing the count of samples by day of week
ggplot(sites,aes(x=dow)) +
  geom_bar() #making a bar plot. 
```



```{r add levels to our data}
#Update the levels to include all seven days of the week (note we have to start with existing levels)
levels(sites$dow) <- c(levels(sites$dow),"1","2","6","7") #this added the numbers / days of the week that were missing

#Confirm the change
levels(sites$dow)

#Plot again
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE) #scales let you override defaults. aka things with values of zero would be dropped

```

We now have all days of the week, but the order is incorrect. Also, we have day numbers, not day names. How to fix?
```{r reorder and rename levels}

#Reorder by overwriting the column with the existing values assigned with a new order of levels
sites$dow <- factor(sites$dow, levels=c("1","2","3","4","5","6","7")) #Or... as.character(seq(1,7)) this changes the order of levels

#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)

#Add **labels** to our factors THIS IS REDOING LABELS FROM LEVELS
sites$dow <- factor(
  sites$dow, 
  levels=c("1","2","3","4","5","6","7"),
  labels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat"))

#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)
```

## An exercise for you: see if you can plot the number of samples recorded each month