---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Claire Pajka"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
install.packages("agricolae")
library(tidyverse);library(lubridate);library(here); library(agricolae); library(dplyr)
here()
chemphysics_raw <- read.csv(
  here("Data", "Raw", "NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE)
View(chemphysics_raw)
chemphysics_raw$sampledate <-
  mdy(chemphysics_raw$sampledate)

#View(chemphysics_raw)
#mutate(sampledate = as.Date(chemphysics_raw$sampledate, format = "%m-%d-%Y"),  # Ensure sampledate is in the correct format
         #Month = month(sampledate),
        #Year = year(sampledate))
#View(chemphysics_raw)

#2
mytheme <- theme_classic(base_size=12)+
  theme(
    plot.title= element_text(size= 12, color = "darkblue", 
                             face="bold", hjust = 0.5),
    axis.text = element_text(size = 12, color = "black"),
    legend.position = 'right',
    legend.background = element_blank(),
    legend.box.background = element_rect(colour = "black"),
    plot.background = element_rect(color = "black"),
    axis.line = element_line(size = 0.65, color = "black"))
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no difference in lake temperatures at different depths across the study lakes.
Ha: Lake temperatures vary by depth across the study lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
subsetdepth_temp <- chemphysics_raw %>%
  #mutate(sampledate = as.Date(sampledate, format = "%m- %d - %Y"),
         #Month = month(sampledate),
         #Year = year(sampledate))%>%
  filter(month(chemphysics_raw$sampledate)== 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()
  view(subsetdepth_temp)

#5
Depth_temp_plot <-
  ggplot(subsetdepth_temp,
         aes(
           y=temperature_C,
           x= depth,
           color = lakename))+
  geom_point()+
  mytheme+
  geom_smooth(method = lm, linewidth = .75, color = "black")+
  xlim(0,20)+
  ylim(0, 35)+
  labs(title= "July Temperature Difference by Depth", 
       fontface = "bold",
       color = "Lake Name")+
  xlab("Depth")+
  ylab("Temperature in Celcius")
print(Depth_temp_plot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
Generally, the graph shows that as depth increases, the temperature in degrees Celcius decreases. The data appears to decrease in a nonlinear manner (as demonstrated by its difference from the trend line), similar to a logarithmic curve. This suggests that the relationship between temperature in Celcius and depth is nonlinear.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
depth_temp_regression <-
  lm(subsetdepth_temp$depth ~ 
       subsetdepth_temp$temperature_C)
summary(depth_temp_regression)

depth_temp_regression_plot <-
  ggplot(subsetdepth_temp, aes(x = depth, y = temperature_C))+
  ylim(0,35)+
  geom_point()
print(depth_temp_regression_plot)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The r^2 explains how much variability is explained by depth, so the r^2 value of 0.7387 means that 73.87% of the variability is accounted for by depth. The degrees of freedom are 9726. The p value is less than 2.2e-16, which is less than 0.05, so we would call this result statistically significant.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TempbyAll.regression <- lm(data = subsetdepth_temp, temperature_C ~ year4 + daynum + depth)
TempbyAll.regression
step(TempbyAll.regression)
print(TempbyAll.regression)
#10
Temp_by_depthmultiple <- lm(data = subsetdepth_temp, temperature_C ~ year4 + daynum + depth)
summary(Temp_by_depthmultiple)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of variables includes year4, daynum, and depth. 
For the multiple regression with the variables suggested by the AIC method, the multiple r squared is 0.7412 (adjusted 0.7411), meaning that 74.12% of the variance is accounted for by the model. The single linear regression of only depth and temperature provided an r squared of 0.7387, which means that the adding in more explanatory variables very slightly increased the amount of the observed variance that the model explains.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
July_temp_anova <- aov(data = subsetdepth_temp, temperature_C ~ lakename)
summary(July_temp_anova)

July_temp_anova2 <- lm(data = subsetdepth_temp, temperature_C ~ lakename)
summary(July_temp_anova2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Given that the p value is less than 0.05, we reject the null hypothesis that there is no significant difference between mean temperature among the lakes. The data implies that there is a significant difference in mean temperature among the lakes.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
Depth_temp_plot2 <-
  ggplot(subsetdepth_temp,
         aes(
           y=temperature_C,
           x= depth,
           color = lakename))+
  geom_point(alpha= 0.5)+
  mytheme+
  geom_smooth(method = lm, se= FALSE, linewidth = .75, color = "black")+
  xlim(0,20)+
  ylim(0, 35)+
  labs(title= "July Temperature Difference by Depth", 
       fontface = "bold",
       color = "Lake Name")+
  xlab("Depth")+
  ylab("Temperature in Celcius")
print(Depth_temp_plot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
Temp.groups <- HSD.test(July_temp_anova, "lakename", group = TRUE)
Temp.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically speaking, Paul Lake has the same mean as Peter Lake. There are no lakes that have a mean temperature that is statistically distinct from all of the other lakes: all of the lakes have at least one other lake with which the means are the same. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
crampton_ward <- subsetdepth_temp %>%
  filter(lakename == "Crampton Lake"| lakename =="Ward Lake")
view(crampton_ward)

july_temp_twosample <- t.test(crampton_ward$temperature_C ~ crampton_ward$lakename)
july_temp_twosample
```

>Answer: Running the two sample t test, we get a p value of 0.2649. Because this is greater than 0.05, we fail to reject the null. The null hypothesis of the two sample t test is that the means are the same, and because we failed to reject the null, the data asserts that the mean July temperatures for Cramptom Lake and Ward Lake are the same. This matches with the answer in #16, because we saw that Crampton Lake had a grouping of ab, and Ward Lake had a grouping of bc. Because they have the group 'b' in common, question 16 showed that there would not be a statistical difference between the means of Crampton and Ward Lake.
