---
title: "Assignment 3: Data Exploration"
author: "Claire Pajka"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Import_Rename_CSV,  message = FALSE}
#imported csv files from file, import dataset, from base text
Neonics <- read.csv(file = "C:/Users/cepaj/OneDrive/Documents/EDE_Fall2023/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv(file = "C:/Users/cepaj/OneDrive/Documents/EDE_Fall2023/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE) #R doesn't like backslashes

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Ecotoxicology refers to the way that substances and chemicals interact and affect environmental systems. Since Neonicotinoids have high toxicity specifically to insects and are used in agriculture, it is helpful to know how they affect different insects. For example, bees and ladybugs are helpful pollinators that are also insects, and are often not the target of most insecticides, which typically aim to kill bugs that feed on the crop plants, like whitebugs. This dives into the debate of "kill the 'bad' bugs, keep the 'good' bugs".

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Nutrient cycling is heavily impacted by the amount of litter and woody debris on the ground, as well as biodiversity of microbes and fungi. Plant productivity can also be measured using woody debris and biomass. Broader implications of studying ground litter and woody debris include assessing a forest's risk for wildfire and creating wildfire management and mitigation plans, such as prescribed burns to decrease fuel for more catastrophic, fast-spreading fires.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter and woody debris are sampled from litter trap pairs (one elevated trap and one ground trap) that are located every 400 square meters of the research area.Litter and woody debris are sorted into separate functional groups after collection, including leaves, needles, twigs/branches, woody material (cones, bark), seeds, flowers, lichen/mosses, and mixed material. The trap areas may be targeted or randomized. 
 2. The trap areas may be targeted or randomized depending on the type of vegetation. Sites with greater than 50% aerial cover with woody vegetation are chosen randomly. In areas with less than 50% woody vegetation cover, the sites are distributed randomly.
 3.Sampling collection frequency depends on the type of forest: ground traps are sampled once per year, elevated traps in deciduous forest sites are sampled once every 2 weeks, and elevated traps at evergreen sites are sampled once every 1-2 months. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dataset_Dimensions}
dim(Neonics) #4623 rows, 30 columns
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Effect_Summary}
summary(Neonics$Effect)
sort(summary(Neonics$Effect), decreasing = TRUE)
```

> Answer: Effects like population, mortality, behavior, and feeding behavior are all extremely important in regards to direct agricultural significance: If a bug is dead then it isn't killing crops (mortality & population effects), whether the bug continues to eat the crop treated with the insecticide (feeding behavior) can tell you the efficacy of the insecticide.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Neonics_Common_Name_Summary}
summary(Neonics$Species.Common.Name)
sort(summary(Neonics$Species.Common.Name), decreasing = TRUE)
```

> Answer: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, Italian Honeybee. These are all species of interest in that they are likely not the intended target of the insecticide: they are all crucial pollinators, and studying adverse affects to these species in response could tell researchers how much 'collateral damage' to unintended species there is from nicotinoid use.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Conc_Author_Class}
class(Neonics$Conc.1..Author.)
```

> Answer: It is a factor class. This may be because certain entries have non-numeric data in them. Some rows have NR in place of a number, Rstudio will not classify NR as a numeric value.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Frequency_Plot}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Neonics$Publication.Year)) #do I need limits here

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Frequency_Plot_Colors}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color=Test.Location), bins=100) +
  scale_x_continuous(limits = c(1982,2014))+
  theme(legend.position = "top")
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Test location does vary over time. Between 1990 and 2000, natural fields were a more common test location than labs, however, after 2000, the lab test locations started increasing and was higher than natural field test locations, especially in the years following 2010. This can be displayed better after increasing the bins from 50 to 100. Artificial fields tended to be lower than both natural fields and lab test locations.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r BarGraph_Endpoint}
ggplot(Neonics, aes(x= Endpoint))+
         geom_bar()
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
sort(summary(Neonics$Endpoint), decreasing = TRUE) #Double checking that graph matches the data shown.
```

> Answer:The two most common endpoints are NOEL and LOEL. According to the EPAs' ECOTOX Code Appendix, NOEL stands for No Observable Effect Level and is defined as "the highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test." LOEL is used for terrestrial responses, stands for Lowest Observable Effect Level, and is defined as "the lowest dose producing effects statistically different from responses of controls."

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r CollectDate_Exploration}
class(Litter$collectDate) #factor originally, not date
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate)
unique(Litter$collectDate) #August 2nd 2018, and August 30th 2018
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Unique_Litter}
unique(Litter$namedLocation)
length(unique(Litter$namedLocation))
summary(Litter$namedLocation)
```

> Answer: 12 plots were sampled. The information given in Summary provides statistical values, so in this case, the number of times litter was sampled at each given location. The unique function is not counting the number of times each plot was sampled, but counting each sample site only once so that the output accounts for every individual (unique) site.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r FunctionalGroup_Bargraph}
ggplot(Litter, aes(x=functionalGroup))+
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Geomboxplotgraph_Geomviolingraph}
ggplot(Litter)+
  geom_boxplot(aes(x=dryMass, y=functionalGroup))
ggplot(Litter)+
  geom_violin(aes(x=dryMass, y=functionalGroup),
               draw_quantiles = c(0.25, 0.5, 0.75))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: There are outliers in the data (especially for twigs/branches, and needles) that can be seen clearly on the boxplot. The violin plot doesn't show this outlier, and displays the drymass as a higher value than it is in reality (it draws the line all the way to the outlier, and doesnt visual the outlier.) OR Because the functional groups were evenly distributed, the violin plot is not as applicable than the box plot.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles and mixed litter tend to have the highest biomass at these sites, followed by twigs/branches with a fairlt lower biomass.
