---
title: "Assignment 2: Coding Basics"
author: "Claire Pajka"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. seq(1,30,3)
sequence_basic_A02 <- seq(1,30,3) #naming the sequence
#2. mean(sequence_basic_A02)
median(sequence_basic_A02)
#3. mean(sequence_basic_A02) > median(sequence_basic_A02) #asking if the mean is greater than the median

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
student_name <- c("Jill","Zach", "Sarah", "Joe") #character data
test_score <- c(90, 45, 80, 60) #double data
pass_test <- test_score >= 50 #logical data
data.frame(student_name, test_score, pass_test)
student_passing <- data.frame(student_name, test_score, pass_test)
colnames(student_passing) <- c("Student", "Score", "Passed")
typeof(student_name)
typeof(test_score)
typeof(pass_test)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame can contain different types of data, like character in column 1, double data in column 2, and logical data in column 3. A matrix can only contain one type of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
